An Improved Approximate-Bayesian Model-choice Method for Estimating Shared Evolutionary History

### 2014-02-19 manuscripts/dpp-msbayes/dpp-msbayes-main.tex
Making some edits to the DPP manuscript.

\subsection{The model}
In this section, I describe the model, which is a modification of the model
implemented in \msb \citep{Huang2011,Oaks2012}.
The code implementing the new model is freely available in the open-source
software package \dppmsbayes
(\href{https://github.com/joaks1/dpp-msbayes}{\url{https://github.com/joaks1/dpp-msbayes}}).
To perform the analyses described below, I used the freely avaliable,
open-source software package \pymsbayes
(\href{https://github.com/joaks1/PyMsBayes}{\url{https://github.com/joaks1/PyMsBayes}}),
which is a Python API that provides a multi-processing interface to \msb and
\dppmsbayes.
I performed the work described below following the principles of
Open Notebook Science.
Using version-control software, I make progress in all aspects of the work
freely and publicly available in real-time at
\href{https://github.com/joaks1/msbayes-experiments}{\url{https://github.com/joaks1/msbayes-experiments}}.
All information necessary to reproduce my results is provided there.
I follow much of the notation of \citet{Oaks2012}, but modify it to
aid in my description of the new model.
A summary of my notation can be found in Table \ref{tabNotation}.

I assume an investigator is interested in inferring the distribution
of divergence times among \npairs{} pairs of populations.
For each pair $i$, \popSampleSize{i}{} genome copies have been sampled, with
\popSampleSize{1}{i} copies sampled from population 1, and \popSampleSize{2}{i}
sampled from population 2.
From these genomes, let \nloci{i} be the number of DNA sequence loci collected
for population pair $i$, and \nlociTotal be the total number of unique loci
sampled across the \npairs{} pairs of populations.
I use \alignment{i}{j} to represent the multiple sequence alignment of
locus $j$ for population pair $i$.
$\alignmentVector = (\alignment{1}{1}, \ldots,
    \alignment{\npairs{}}{\nloci{\npairs{}}})$
is the full dataset,
i.e., a vector of sequence alignments for all pairs and loci.
I let \geneTree{i}{j} represent the gene tree upon which \alignment{i}{j}
evolved according to fixed \hky substitution model parameters \hkyModel{i}{j}.
The investigator must specify the parameters of all
$\hkyModelVector = (\hkyModel{1}{1}, \ldots,
\hkyModel{\npairs{}}{\nloci{\npairs{}}})$
substitution models by which the alignments evolved along the
$\geneTreeVector = (\geneTree{1}{1}, \ldots,
\geneTree{\npairs{}}{\nloci{\npairs{}}})$
gene trees.
Furthermore, the investigator must specify a vector of fixed constants
$\ploidyScalarVector = (\ploidyScalar{1}{1}, \ldots,
\ploidyScalar{\npairs{}}{\nloci{\npairs{}}})$
that scale the population-size parameters for known differences in ploidy among
loci and/or differences in generation times among population pairs.  Lastly,
the investigator must also specify a vector of fixed constants
$\mutationRateScalarConstantVector = (\mutationRateScalarConstant{1}{1},
\ldots, \mutationRateScalarConstant{\npairs{}}{\nloci{\npairs{}}})$
that scale the population-size parameters for known differences in
mutation rates among loci and/or among taxa.

@@

\subsubsection{Priors on divergence models}
\begin{linenomath}
The prior term for the vector of divergence times for each of
the \npairs{} pairs of populations can be expanded as
\begin{equation}
    p(\divTimeMapVector) = p(\divTimeIndexVector)p(\divTimeVector \given \divTimeIndexVector),
    \label{eq:divModelPrior}
\end{equation}
where \divTimeVector is an ordered set of unique divergence-time parameters
$\{\divTime{1}, \ldots, \divTime{\divTimeNum}\}$ whose length
\divTimeNum can range from 1 to \npairs{},
and \divTimeIndexVector is the vector of indices of divergence-time parameters 
$(\divTimeIndex{1}, \ldots, \divTimeIndex{\npairs{}})$
that map the divergence times in \divTimeVector to each of the \npairs{}
pairs of populations.
Thus, \divTimeMapVector is the result of applying the mapping function
\begin{equation}
    f(\divTimeVector, \divTimeIndexVector, i) = \divTime{\divTimeIndex{i}}
    \label{eq:divTimeMapFunction}
\end{equation}
to each population pair $i$, such that
$\divTimeMapVector = (\divTimeMap{1} = f(\divTimeVector, \divTimeIndexVector,
1), \ldots, \divTimeMap{\npairs{}} = f(\divTimeVector, \divTimeIndexVector,
\npairs{}))$.
\end{linenomath}

Biologically speaking, \divTimeVector contains the times of divergence events,
the length of which \divTimeNum is the number of divergence events shared across
the \npairs{} pairs of populations.
For example, if \divTimeVector contains a single divergence time parameter
\divTime{1}, all \npairs{} pairs of populations are constrained to diverge at
this time (i.e., \divTimeIndexVector would contain the index 1 repeated
\npairs{} times, and \divTimeMapVector would contain the value \divTime{1}
repeated \npairs{} times), whereas if it contains \npairs{} divergence time
parameters, the model is fully generalized to allow all of the pairs to diverge
at unique times.

>>>>

\subsection{The model}
In this section, I describe the model, which is a modification of the model
implemented in \msb \citep{Huang2011,Oaks2012}.
The code implementing the new model is freely available in the open-source
software package \dppmsbayes
(\href{https://github.com/joaks1/dpp-msbayes}{\url{https://github.com/joaks1/dpp-msbayes}}).
To perform the analyses described below, I used the freely avaliable,
open-source software package \pymsbayes
(\href{https://github.com/joaks1/PyMsBayes}{\url{https://github.com/joaks1/PyMsBayes}}),
which is a Python API that provides a multi-processing interface to \msb and
\dppmsbayes.
I performed the work described below following the principles of
Open Notebook Science.
Using version-control software, I make progress in all aspects of the work
freely and publicly available in real-time at
\href{https://github.com/joaks1/msbayes-experiments}{\url{https://github.com/joaks1/msbayes-experiments}}.
All information necessary to reproduce my results is provided there.
I follow much of the notation of \citet{Oaks2012}, but modify it to
aid in the description of the new model.
A summary of my notation can be found in Table \ref{tabNotation}.

I assume an investigator is interested in inferring the distribution
of divergence times among \npairs{} pairs of populations.
For each pair $i$, \popSampleSize{i}{} genome copies have been sampled, with
\popSampleSize{1}{i} copies sampled from population 1, and \popSampleSize{2}{i}
sampled from population 2.
From these genomes, let \nloci{i} be the number of DNA sequence loci collected
for population pair $i$, and \nlociTotal be the total number of unique loci
sampled across the \npairs{} pairs of populations.
I use \alignment{i}{j} to represent the multiple sequence alignment of
locus $j$ for population pair $i$.
$\alignmentVector = (\alignment{1}{1}, \ldots,
    \alignment{\npairs{}}{\nloci{\npairs{}}})$
is the full dataset,
i.e., a vector of sequence alignments for all pairs and loci.
Let \geneTree{i}{j} represent the gene tree upon which \alignment{i}{j}
evolved according to fixed \hky substitution model parameters \hkyModel{i}{j}.
The investigator must specify the parameters of all
$\hkyModelVector = (\hkyModel{1}{1}, \ldots,
\hkyModel{\npairs{}}{\nloci{\npairs{}}})$
substitution models by which the alignments evolved along the
$\geneTreeVector = (\geneTree{1}{1}, \ldots,
\geneTree{\npairs{}}{\nloci{\npairs{}}})$
gene trees.
Furthermore, the investigator must specify a vector of fixed constants
$\ploidyScalarVector = (\ploidyScalar{1}{1}, \ldots,
\ploidyScalar{\npairs{}}{\nloci{\npairs{}}})$
that scale the population-size parameters for known differences in ploidy among
loci and/or differences in generation times among population pairs.  Lastly,
the investigator must also specify a vector of fixed constants
$\mutationRateScalarConstantVector = (\mutationRateScalarConstant{1}{1},
\ldots, \mutationRateScalarConstant{\npairs{}}{\nloci{\npairs{}}})$
that scale the population-size parameters for known differences in
mutation rates among loci and/or among taxa.

@@

\subsubsection{Priors on divergence models}
\begin{linenomath}
The prior term for the vector of divergence times for each of the \npairs{}
pairs of populations, \divTimeMapVector, can be expanded as
\begin{equation}
    p(\divTimeMapVector) = p(\divTimeIndexVector)p(\divTimeVector \given \divTimeIndexVector),
    \label{eq:divModelPrior}
\end{equation}
where \divTimeVector is an ordered set of divergence-time parameters
$\{\divTime{1}, \ldots, \divTime{\divTimeNum}\}$ whose length
\divTimeNum can range from 1 to \npairs{},
and \divTimeIndexVector is a vector of indices %of divergence-time parameters 
$(\divTimeIndex{1}, \ldots, \divTimeIndex{\npairs{}})$,
where
$\divTimeIndex{i} \in \{1, \ldots, \divTimeNum\}$.
These indices map each of the \npairs{} pairs of populations to a
divergence-time parameter in \divTimeVector.
Thus, \divTimeMapVector is the result of applying the mapping function
\begin{equation}
    f(\divTimeVector, \divTimeIndexVector, i) = \divTime{\divTimeIndex{i}}
    \label{eq:divTimeMapFunction}
\end{equation}
to each population pair $i$, such that
$\divTimeMapVector = (\divTimeMap{1} = f(\divTimeVector, \divTimeIndexVector,
1), \ldots, \divTimeMap{\npairs{}} = f(\divTimeVector, \divTimeIndexVector,
\npairs{}))$.
\end{linenomath}

Biologically speaking, \divTimeVector contains the times of divergence events,
the length of which \divTimeNum is the number of divergence events shared across
the \npairs{} pairs of populations.
For example, if \divTimeVector contains a single divergence-time parameter
\divTime{1}, all \npairs{} pairs of populations are constrained to diverge at
this time (i.e., \divTimeIndexVector would contain the index 1 repeated
\npairs{} times, and \divTimeMapVector would contain the value \divTime{1}
repeated \npairs{} times), whereas if it contains \npairs{} divergence-time
parameters, the model is fully generalized to allow all of the pairs to diverge
at unique times.

### 2014-02-19 manuscripts/dpp-msbayes/dpp-msbayes.tex
Updating author line of dpp manuscript.
Making some edits to the DPP manuscript.

\author[1]{Jamie R.\ Oaks\thanks{Corresponding author: \href{mailto:joaks1@gmail.com}{\tt joaks1@gmail.com}}\thanks{Current address: Department of Biology, University of Washington, Seattle, Washington 98195}}
\author[1]{Mark T.\ Holder}
\affil[1]{Department of Ecology and Evolutionary Biology, University of Kansas, Lawrence, Kansas 66045}

@@

\begin{abstract}
    To understand the processes that generate biodiversity, it is
    important to account for large-scale processes that affect the evolutionary
    history of groups of co-distributed populations of organisms.
    Such events predict temporally clustered divergences times, a pattern
    that can be estimated using genetic data from co-distributed species.
    I introduce a new approximate-Bayesian model for comparative
    phylogeographical model-choice that estimates the temporal distribution of
    divergences across taxa from multi-locus DNA sequence data.
    The model is an extension of that implemented in \msb.
    By reparameterizing the model, introducing more flexible priors on
    demographic and divergence-time parameters, and implementing a
    non-parametric Dirichlet-process prior over divergence models, I improve
    the robustness, accuracy, and power of the method for estimating shared
    evolutionary history across taxa.
    My results demonstrate the improved performance of the new method is due
    to (1) more appropriate priors on divergence-time and demographic
    parameters that avoid prohibitively small marginal likelihoods for models
    with more divergence events,
    and (2) the Dirichlet-process providing a flexible prior on divergence
    histories that does not strongly disfavor models with intermediate numbers
    of divergence events.
    % The results demonstrate that the tendency of the \msb model to spuriously
    % support models of clustered divergences is caused by a combination of (1)
    % uniform priors on nuisance parameters hindering the marginal likelihoods of
    % models with more divergence-time parameters, and (2) a prior on divergence
    % models that disfavors models with intermediate numbers of divergence-time
    % parameters.
    The new method yields more robust estimates of posterior uncertainty, and
    thus greatly reduces the tendency of the model to incorrectly estimate
    biogeographically interesting models with strong support.

@@

\section*{Supporting Information}
% PUT MAIN TEXT CITATION HERE
\hangindent=1cm
Oaks, J.~R.\ and M.~T.\ Holder. \msTitle.

>>>>

\author[1]{Jamie R.\ Oaks\thanks{Corresponding author: \href{mailto:joaks1@gmail.com}{\tt joaks1@gmail.com}}\thanks{Current address: Department of Biology, University of Washington, Seattle, Washington 98195}}
% \author[1]{Mark T.\ Holder}
\affil[1]{Department of Ecology and Evolutionary Biology, University of Kansas, Lawrence, Kansas 66045}

@@

\begin{abstract}
    To understand the processes that generate biodiversity, it is
    important to account for large-scale processes that affect the evolutionary
    history of groups of co-distributed populations of organisms.
    Such events predict temporally clustered divergences times, a pattern
    that can be estimated using genetic data from co-distributed species.
    I introduce a new approximate-Bayesian model for comparative
    phylogeographical model-choice that estimates the temporal distribution of
    divergences across taxa from multi-locus DNA sequence data.
    The model is an extension of that implemented in \msb.
    By reparameterizing the model, introducing more flexible priors on
    demographic and divergence-time parameters, and implementing a
    non-parametric Dirichlet-process prior over divergence models, I improved
    the robustness, accuracy, and power of the method for estimating shared
    evolutionary history across taxa.
    The results demonstrate the improved performance of the new method is due
    to (1) more appropriate priors on divergence-time and demographic
    parameters that avoid prohibitively small marginal likelihoods for models
    with more divergence events,
    and (2) the Dirichlet-process providing a flexible prior on divergence
    histories that does not strongly disfavor models with intermediate numbers
    of divergence events.
    % The results demonstrate that the tendency of the \msb model to spuriously
    % support models of clustered divergences is caused by a combination of (1)
    % uniform priors on nuisance parameters hindering the marginal likelihoods of
    % models with more divergence-time parameters, and (2) a prior on divergence
    % models that disfavors models with intermediate numbers of divergence-time
    % parameters.
    The new method yields more robust estimates of posterior uncertainty, and
    thus greatly reduces the tendency of the model to incorrectly estimate
    biogeographically interesting models with strong support.

@@

\section*{Supporting Information}
% PUT MAIN TEXT CITATION HERE
\hangindent=1cm
Oaks, J.\ R. \msTitle.
